---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Introduction
The database contains a large amount of information on personal movement using activity tracking devices. These devices collect data in 5-minute intervals throughout the day from an anonymous individual during the October and November 2012 months and include the number of steps taken in 5-minute intervals each day.

## Enviroment preparation 
```{r}
## Clean the enviroment
rm(list=ls())

## Load libraries
library(tidyverse)
```


## Loading and preprocessing the data
```{r}
## Load the data
db <- read_csv("../data/activity.zip")

db%>%glimpse

db%>%
          summarise_all(~sum(is.na(.)))

```

We can observe that database contains only 3 variables and 17 568 observations. But `steps` variable contains a large number of NA's. 


## What is mean total number of steps taken per day?
```{r}
## Calculate total number of steps taken per day
total_steps <- db%>%
          drop_na()%>%
          group_by(date)%>%
          summarise(sum_steps=sum(steps, na.rm=TRUE))%>%
          print()


## Make a histogram of the total number of steps taken each day
with(total_steps, hist(sum_steps, col="green", 
               main="Total number steps per day",
               xlab="Steps in a day"))


## Calculate and report the mean and median of the total number of steps taken per day
          


```
```{r}

with(db, hist(steps~date))


```


## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
